Reproducing "Fluctuation Domains

### 2020-02-18 rescience/article.Rmd

In this analyis I seek to reproduce the numerical results of [@Boettiger2010], which focuses on the evolutionary trajectory of species trait under selection, as predicted by the so-called "Canonical Equation" of adaptive dynamics [@Dieckmann1996].  Theoretical work I present in the paper predicts the emergence of a 'flutation enhancement' regime far from an evolutionary stable point, and suggests that the evolution of traits which start in this distant region would deviate substantially from the expected evolutionary path predicted by the canonical equation.  In particular, the distribution of such trajectories could be bimodal, making the "canonical" path far from the most likely path.  The details of this distribution cannot be solved for analytically however, so in this paper I resorted to an replicate exact stochastic simulations of the evolutionary model using a Gillespie algorithm [@Gillespie].

@@

The published paper itself states that the code is available at the personal website of one of my co-authors, with URL <http://ecotheory.biology.gatech.edu/downloads/>.  Fortunately, this URL still resolves (my colleague is still at the same institution), and does indeed contain a downloadable zip file with a copy of the code. In 2012, I had uploaded a copy of this code as a supplement to Dryad, but both on my CV a the time and on my colleague's website we had recorded only the Dryad handle URL, <http://datadryad.org/handle/10255/dryad.37625>, which no longer resolves.  That handle can be resolved by the WayBackMachine, which doesn't contain the code archive but does reveal the DOI, <http://dx.doi.org/10.5061/dryad.j8n0p7vc>, which successfully resolves to the Dryad copy.  The code can also be found by searching the Dryad web interface for my surname (though it is listed as "data" and not "software."  In 2017, I had also imported the GitHub copy into the Zenodo data repository where it was again assigned a DOI and is identified as software in the metadata.  I note that the package was never submitted to the Central R Archive Network, CRAN.

@@

>  In the spirit of scientifically reproducible research(Gentleman and Lang, 2007; Schwabet al., 2000; Stodden, 2009), we freely provide all the source code required to replicate the simulations and figuress hown in the text. Though the numerical simulations are written in C for computational efficiency, we provide a user interface and documentation by releasing all the code, figures, text, and examples as a software package for the widely used and freely available R statistical computing language. This package can be downloaded from http://ecotheory.biology.gatech.edu/downloads/.

@@

The same stategy worked on MacOSX (Catalina) with `gsl` installed via Homebrew. I was able to update the original autoconf script by recalling that it had been adapted from another package, the `gsl` R package <https://cran.r-project.org/web/packages/gsl/>, which also binds against the GSL library. Working from the autoconf files provided there, I could successfully get the package to configure and compile.  This highlights a potential risk in changes that are intended to make the package more portable but also add complexity.  More generally, while the package has no dependency on external R packages (beyond those default packages provided as part of base R: `stats`, `utils`, `graphics`, and `grDevices`, this issue illustrates the familiar reproducibility challenge that is created by changing versions of dependencies and system libraries (GSL in this case).  System libraries against which a package is compiled are particularly pernicious, because versioning of these components is both platform dependent and separate from the R package manager process, and thus cannot be addressed merely with package version tracking in tools such as `packrat` [@packrat] or `renv` [@renv].  Today, containerized approaches such a Docker [see @Boettiger2015; @Boettiger2017] provide a solution to this system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the mansucript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@knitr_book]).  In that version, files `fig1.R` and `fig2.R` are were included in the same subdirectory as the `pdf` figures referenced by the `tex` version of the manuscript, but `fig2.R`, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. That result data files had also not been archived seperately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

Running the `logistic()` function provided in the code, I was able to replicate the key results of the paper.  A few qualifications here are necessary.  The paper consists of two core results -- a  theoretical (i.e. derived analytically without computational component) which predicts the existance of 'fluctuation enhancement regimes' which can create substantial deviations from the dynamics predicted by the accepted theory at the time (that is, from the so-called canonical equation of adaptive dynamics), and a second result which relies on a computationally intensive stochastic ensemble simulation to illustrate these deviations arising for dynamics that begin in the "fluctation enhancement" regime.  These results are summarized by figures 1 and 2 of the original paper, which I reproduce here.

@@

```{r fig1, fig.cap="The three panels of figure 1, showing examples of evolutionary landscapes and the corresponding fluctation domains for three classic models (logistic, chemostat, branching model). This figure is largely theoretical, and so easily replicates Figure 1 of the original paper.  The first panel showing the logistic model depicts the fluctation regimes for the model parameters used in the numerical analysis."}
library(fluctuationDomains)
## Figure 1 function definitions:
logistic_curve = function(x){-x*exp(-x*x/2)}
chemostat_curve <- function(x, d = 0.1, q = 0.1){
  (q*x-d/(1-x*d))* (d/x-2*d**2)
}
branching_curve <- function(x, SIGMA2_K = 2, SIGMA2_C = 1){
  -exp( (x^2/2)*(4/SIGMA2_C -1/SIGMA2_K) ) *
    (SIGMA2_C+exp(2*x^2/SIGMA2_C)*SIGMA2_C-2*SIGMA2_K)*x /
    ( (1+exp(2*x^2/SIGMA2_C)^2)*SIGMA2_C*SIGMA2_K)
}

@@

A dynamic document using RMarkdown [@rmarkdown] for reproducing the present mansucript can be found in the `rescience` sub-directory which I have now added to the GitHub repository housing the original R package, <https://github.com/cboettig/fluctuationDomains>, git tag `rescience-submission`, now archived with DOI <https://doi.org/xxxx>.

>>>>

In this analsyis I seek to reproduce the numerical results of [@Boettiger2010], which focuses on the evolutionary trajectory of species trait under selection, as predicted by the so-called "Canonical Equation" of adaptive dynamics [@Dieckmann1996].  Theoretical work I present in the paper predicts the emergence of a 'flutation enhancement' regime far from an evolutionary stable point, and suggests that the evolution of traits which start in this distant region would deviate substantially from the expected evolutionary path predicted by the canonical equation.  In particular, the distribution of such trajectories could be bimodal, making the "canonical" path far from the most likely path.  The details of this distribution cannot be solved for analytically however, so in this paper I resorted to an replicate exact stochastic simulations of the evolutionary model using a Gillespie algorithm [@Gillespie].

@@

The published paper itself states that the code is available at the personal website of one of my co-authors, with URL <http://ecotheory.biology.gatech.edu/downloads/>.  Fortunately, this URL still resolves (my colleague is still at the same institution), and does indeed contain a downloadable zip file with a copy of the code. In 2012, I had uploaded a copy of this code as a supplement to Dryad, but both on my CV at the time and on my colleague's website we had recorded only the Dryad handle URL, <http://datadryad.org/handle/10255/dryad.37625>, which no longer resolves.  That handle can be resolved by the WayBackMachine, which doesn't contain the code archive but does reveal the DOI, <http://dx.doi.org/10.5061/dryad.j8n0p7vc>, which successfully resolves to the Dryad copy.  The code can also be found by searching the Dryad web interface for my surname (though it is listed as "data" and not "software."  In 2017, I had also imported the GitHub copy into the Zenodo data repository where it was again assigned a DOI and is identified as software in the metadata.  I note that the package was never submitted to the Central R Archive Network, CRAN.

@@

>  In the spirit of scientifically reproducible research(Gentleman and Lang, 2007; Schwabet al., 2000; Stodden, 2009), we freely provide all the source code required to replicate the simulations and figures shown in the text. Though the numerical simulations are written in C for computational efficiency, we provide a user interface and documentation by releasing all the code, figures, text, and examples as a software package for the widely used and freely available R statistical computing language. This package can be downloaded from http://ecotheory.biology.gatech.edu/downloads/.

@@

The same strategy worked on MacOSX (Catalina) with `gsl` installed via Homebrew. I was able to update the original autoconf script by recalling that it had been adapted from another package, the `gsl` R package <https://cran.r-project.org/web/packages/gsl/>, which also binds against the GSL library. Working from the autoconf files provided there, I could successfully get the package to configure and compile.  This highlights a potential risk in changes that are intended to make the package more portable but also add complexity.  More generally, while the package has no dependency on external R packages (beyond those default packages provided as part of base R: `stats`, `utils`, `graphics`, and `grDevices`, this issue illustrates the familiar reproducibility challenge that is created by changing versions of dependencies and system libraries (GSL in this case).  System libraries against which a package is compiled are particularly pernicious, because versioning of these components is both platform dependent and separate from the R package manager process, and thus cannot be addressed merely with package version tracking in tools such as `packrat` [@packrat] or `renv` [@renv].  Today, containerized approaches such as Docker [see @Boettiger2015; @Boettiger2017] provide a solution to this system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the manuscript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@knitr_book]).  In that version, files `fig1.R` and `fig2.R` are were included in the same subdirectory as the `pdf` figures referenced by the `tex` version of the manuscript, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. The result data files had also not been archived separately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

Running the `logistic()` function provided in the code, I was able to replicate the key results of the paper.  A few qualifications here are necessary.  The paper consists of two core results -- a  theoretical (i.e. derived analytically without computational component) which predicts the existence of 'fluctuation enhancement regimes' which can create substantial deviations from the dynamics predicted by the accepted theory at the time (that is, from the so-called canonical equation of adaptive dynamics), and a second result which relies on a computationally intensive stochastic ensemble simulation to illustrate these deviations arising for dynamics that begin in the "fluctuation enhancement" regime.  These results are summarized by figures 1 and 2 of the original paper, which I reproduce here.

@@

```{r fig1, fig.cap="The three panels of figure 1, showing examples of evolutionary landscapes and the corresponding fluctuation domains for three classic models (logistic, chemostat, branching model). This figure is largely theoretical, and so easily replicates Figure 1 of the original paper.  The first panel showing the logistic model depicts the fluctuation regimes for the model parameters used in the numerical analysis."}
library(fluctuationDomains)
## Figure 1 function definitions:
logistic_curve = function(x){-x*exp(-x*x/2)}
chemostat_curve <- function(x, d = 0.1, q = 0.1){
  (q*x-d/(1-x*d))* (d/x-2*d**2)
}
branching_curve <- function(x, SIGMA2_K = 2, SIGMA2_C = 1){
  -exp( (x^2/2)*(4/SIGMA2_C -1/SIGMA2_K) ) *
    (SIGMA2_C+exp(2*x^2/SIGMA2_C)*SIGMA2_C-2*SIGMA2_K)*x /
    ( (1+exp(2*x^2/SIGMA2_C)^2)*SIGMA2_C*SIGMA2_K)
}

@@

A dynamic document using RMarkdown [@rmarkdown] for reproducing the present manuscript can be found in the `rescience` sub-directory which I have now added to the GitHub repository housing the original R package, <https://github.com/cboettig/fluctuationDomains>, git tag `rescience-submission`, now archived with DOI <https://doi.org/xxxx>.
